---
title: "Global biodiversity of marine mammals"
output: html_document
---

```{r setup, include=FALSE}
library(ape)
library(fasterize)
library(picante)
library(raster)
library(rotl)
library(sf)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

*CSUMB REU 2020*  
*Introduction to Coding for Emerging Geoscientists Workshop*  
*August 5, 2020*

## Introduction

In this part of the workshop, we will use computational tools for a geospatial
analysis of marine mammal biodiversity. Drawing on the concepts of phylogeny and
thermal physiology, we will map two different metrics of marine mammal 
biodiversity to test the metabolic asymmetry hypothesis. For more details, see:

Grady, J. M., Maitner, B. S., Winter, A. S., Kaschner, K., Tittensor, D. P., 
Record, S., ... & Wearing, H. J. (2019). Metabolic asymmetry and the global 
diversity of marine predators. *Science*, 363(6425).

## Loading data

The data for this analysis are all publicly available: 

  * Species range maps: International Union for Conservation of Nature (IUCN)
  www.iucnredlist.org
  * Phylogeny: Upham, N. S., Esselstyn, J. A., & Jetz, W. (2019). Inferring the 
  mammal tree: Species-level sets of phylogenies for questions in ecology, 
  evolution, and conservation. *PLOS Biology*, 17(12), e3000494.
  * Coastlines: Natural Earth www.naturalearthdata.com
  
The following chunk loads each of the three data sources. You should see a range
map of the northern bottlenose whale after running it.

```{r load}
# Ranges
marmam_shp <- st_read("data/marmam.shp")
# Phylogeny
marmam_tr <- read.tree("data/marmam.nex")
# Coastlines
coastline_shp <- st_read("data/coastline.shp")

# Range of the northern bottlenose whale
ggplot(filter(marmam_shp, BINOMIAL == "Hyperoodon ampullatus")) +
  geom_sf(fill = "darkblue") +
  geom_sf(data = coastline_shp) +
  theme_void()
```

**Exercise 1**: make a range map for Hector's dolphin.

## Spatial data types

Spatial data can be represented in two ways: as *vectors* (points, lines, and
polygons) or *rasters* (a rectangular grid of cells). Vectors can represent fine
details better than rasters, but most computational analyses require data in
raster form. Here we're going to convert the vector representation of all the
marine mammal ranges to a *raster brick*, which is basically a bunch of rasters 
with the same grid stacked on top of each other.

When dealing with rasters, we will *project* our spatial data. Since maps are 2D
but the earth is 3D, all maps approximations that warp and twist a sphere onto a
flat piece of paper. Don't worry about the details. If you take a GIS class you
will learn all about it.

```{r raster}
# We'll use the Behrmann equal area projection for all our rasters
cea_proj4 <- "+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=km +no_defs"

# Then we create a raster template to define extent and resolution
template <- raster(
  crs = CRS("+proj=longlat +datum=WGS84"),
  xmn = -180, xmx = 180,
  ymn = -90, ymx = 90
) %>% 
  { suppressWarnings(projectRaster(., crs = cea_proj4)) }
res(template) <- 110

# Finally we rasterize the marine mammal ranges. The resulting brick has a layer 
# for each species.
marmam_rb <- fasterize(
  st_transform(marmam_shp, cea_proj4),
  template,
  by = "BINOMIAL",
  background = 0
) 
names(marmam_rb) <- str_replace(names(marmam_rb), "\\.", "_")

# Plot the raster representation of the northern bottlenose whale range
nbw_rng <- marmam_rb[[which(names(marmam_rb) == "Hyperoodon_ampullatus")]]
ggplot() +
  geom_sf(data = st_transform(coastline_shp, cea_proj4)) +
  geom_raster(aes(x, y, alpha = Hyperoodon_ampullatus == 1),
              as.data.frame(nbw_rng, xy = TRUE),
              fill = "darkblue") +
  scale_alpha_manual(values = c(0, 1)) +
  theme_void() +
  theme(legend.position = "none")
```

**Excerise 2**: how big are the raster cells and how many are there? Hint: use
the `res` and `ncell` functions at the console.

## Alpha diversity

Now that we have our range data in raster format, we're ready for computational
methods. In this case, we want to know the *alpha diversity* of marine mammals.
If you recall, alpha diversity is the number of species in a community. We'll
treat each grid cell as a community and count the overlapping ranges.

**Exercise 3**: fill in "???" with the correct variable. What's the highest 
alpha diversity value and (roughly) where is it located?

```{r alpha}
marmam_alpha <- sum(???, na.rm = TRUE)
plot(marmam_alpha)
```

## Phylogenetic diversity

As discussed earlier, alpha diversity is not the only biodiversity metric. 
Faith's phylogenetic diversity (PD) is correlated with alpha diversity, but it 
better represents the evolutionary relationships in communities. We're going to
use a phylogenetic tree to calculate PD.

```{r phylo}
marmam_tr <- read.tree("data/marmam.nex")
tree_from_xy <- function(x, y) {
  pt <- data.frame(x = x, y = y) %>% 
    st_as_sf(coords = 1:2, crs = 4326) %>% 
    st_transform(cea_proj4) %>% 
    st_coordinates()
  sp <- names(marmam_rb)[extract(marmam_rb, pt) == 1]
  n_sp <- length(sp)
  marmam_tr %>% 
    keep.tip(sp) %>% 
    plot(main = sprintf("(%.1f\u00B0, %.1f\u00B0)", y, x),
         sub = sprintf("%d species", n_sp))
}
```

**Exercise 4**: use the `tree_from_xy` function to figure out how many cetacean 
and pinniped species are found in these locations:

  * 100 miles west of Montery
  * The Mediterranean Sea
  * Hudson Bay
  
With this tree, we can now calculate and map PD of marine mammals globally. 
Expect this chunk to run for a longer time.

```{r pd}
marmam_mtx <- map(seq(nlayers(marmam_rb)), ~ as.vector(marmam_rb[[.x]])) %>% 
  do.call(cbind, .)
colnames(marmam_mtx) <- names(marmam_rb)
rownames(marmam_mtx) <- seq(ncell(marmam_rb))
empty_rows <- apply(marmam_mtx, 1, function(row) all(is.na(row)))
marmam_pd_tbl <- pd(marmam_mtx[!empty_rows, ], marmam_tr) %>% 
  as_tibble(rownames = "i") %>% 
  mutate(i = as.numeric(i)) %>% 
  complete(i = seq(ncell(marmam_rb)))
marmam_pd_r <- setValues(template, marmam_pd_tbl$PD)
marmam_pd_tbl <- marmam_pd_r %>% 
  as.data.frame(xy = TRUE) %>% 
  rename(pd = layer)

ggplot() +
  geom_raster(aes(x, y, fill = pd, alpha = pd == 0), marmam_pd_tbl) +
  geom_sf(data = coastline_shp) +
  scale_fill_gradientn("PD",
                       colors = c("dark blue", "light blue", "pink", "red")) +
  scale_alpha_manual(values = c(1, 0), guide = FALSE) +
  theme_void() +
  theme(legend.direction = "horizontal",
        legend.position = "bottom")
```

**Exercise 5**: approximately how long did it take to calculate PD manually for 
one community? If oceans cover 70% of the earth, how long would it take to 
calculate PD by hand for every oceanic grid cell?

**Exercise 6**: where is marine mammal PD highest? How does that compare to 
alpha diversity?

**Exercise 7**: the `tree_from_xy` function again, investigate the difference
between PD and alpha diversity. What pattern do you see in regions with high
alpha diversity and low PD? How about in regions with low alpha/high PD? You'll
probably have to google the common names of species to figure this out.

**Exercise 8**: how would the metabolic asymmetry hypothesis explain the 
global patterns in marine mammal PD? How would you expect a map of predatory 
fish PD to look different from the marine mammal map?

**Optional exercise**: all the data needed to map the phylogenetic diversity of
fish is included in this project. Using this .Rmd file as a guide, create the
PD map for fish and compare it to your predictions from exercise 8.

